---
title: "Architecture of the DuckDBDataFrame Package"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r format(Sys.time(), '%B %d, %Y')`"
package: DuckDBDataFrame
vignette: >
  %\VignetteIndexEntry{Architecture of the DuckDBDataFrame Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 4
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("DuckDBDataFrame")` package provides DuckDB-backed implementations
of core tabular data structures for the Bioconductor project. It serves as the
foundational layer for the DuckDB-backed Bioconductor ecosystem, which includes
`r Biocpkg("DuckDBArray")` for array/matrix operations and 
`r Biocpkg("DuckDBGRanges")` for genomic interval operations.

## Position in the Bioconductor Ecosystem

`r Biocpkg("DuckDBDataFrame")` extends the `r Biocpkg("S4Vectors")` framework,
implementing DuckDB-backed versions of core classes:

| Class | Extends | Purpose |
|-------|---------|---------|
| *DuckDBTable* | *RectangularData* | N-dimensional table with SQL backend |
| *DuckDBDataFrame* | *DataFrame* | 2D tabular data with row/column names |
| *DuckDBColumn* | *Vector* | Single column extraction |
| *DuckDBAtomicList* | *List* | List columns (DuckDB LIST[]) |
| *DuckDBEmbeddings* | *Matrix* | Fixed-length arrays (DuckDB ARRAY[]) |

The package is designed as an **infrastructure package** for developers building
DuckDB-backed Bioconductor tools, while remaining accessible to end users for
large-scale data manipulation.

## Installation

```{r install, eval=FALSE}
if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("DuckDBDataFrame")
```

```{r load-package}
library(DuckDBDataFrame)
```


# Core Design Principles

## The N-Dimensional Table Abstraction

At the heart of `r Biocpkg("DuckDBDataFrame")` is *DuckDBTable*, an
n-dimensional table class that serves as the backend for both 2D DataFrames
and higher-dimensional arrays. This design choice enables:

1. **Unified SQL operations**: All operations translate to SQL regardless of dimensionality
2. **Flexible key dimensions**: Support for 0, 1, 2, or more key dimensions
3. **Efficient storage**: Parquet's columnar format for all data types
4. **Lazy evaluation**: Operations build SQL queries without immediate execution

### Key Concepts

**keycols**
: Named list defining dimension indices. Each element is an atomic vector
  representing the values along that dimension. Can be integer indices or
  named character/factor levels.

**datacols**
: Named expression defining the data columns. Each element is an R expression
  that translates to a SQL column reference or computation.

**conn**
: A `tbl_duckdb_connection` object from `r CRANpkg("dplyr")` and 
  `r CRANpkg("dbplyr")`, providing the SQL interface to DuckDB.

**dimtbls**
: Optional dimension lookup tables for partitioned data, enabling efficient
  filtering across grid partitions.


## SQL Translation Philosophy

`r Biocpkg("DuckDBDataFrame")` translates R operations into SQL queries that
execute in DuckDB's optimized analytical engine. This provides:

- **Columnar operations**: DuckDB's columnar engine excels at aggregations
- **Lazy evaluation**: Build complex queries without intermediate materialization
- **Memory efficiency**: Process data larger than RAM
- **Parquet optimization**: Direct Parquet file scanning with predicate pushdown


# The DuckDBTable Class

## Class Structure

*DuckDBTable* is the foundational class, extending *RectangularData* from
`r Biocpkg("S4Vectors")`:

```{r DuckDBTable-class}
showClass("DuckDBTable")
```

### Slots

**conn** (`tbl_duckdb_connection`)
: The dplyr/dbplyr connection to a DuckDB table or query. Can reference
  Parquet files, CSV files, or in-memory tables.

**datacols** (`expression`)
: Named expression where each element defines a data column. Elements can be
  simple column references (e.g., `as.name("mpg")`) or complex SQL expressions
  (e.g., `call("/", as.name("mpg"), as.name("hp"))`).

**keycols** (`list`)
: Named list defining dimension indices. For a 2D table with row names, this
  would be a single-element list. For higher dimensions, multiple elements.
  Special encoding for row numbers: `list(row_number = c(NA_integer64_, -n))`.

**dimtbls** (`environment`)
: Locked environment containing a `DataFrameList` of dimension lookup tables.
  Used for partitioned arrays to map dimension indices to grid groups.


## Construction Patterns

### From Parquet Files

The most common construction pattern reads Parquet files directly:

```{r DuckDBTable-from-parquet}
# Write mtcars to Parquet
mtcars_path <- tempfile(fileext = ".parquet")
mtcars_df <- cbind(model = rownames(mtcars), mtcars)
arrow::write_parquet(mtcars_df, mtcars_path)

# Create DuckDBTable with row names
tbl <- DuckDBTable(mtcars_path,
                   datacols = colnames(mtcars),
                   keycols = list(model = mtcars_df$model))
tbl
```

### From CSV Files

DuckDB can read CSV files directly, including gzipped files:

```{r DuckDBTable-from-csv}
# Write to gzipped CSV
csv_path <- tempfile(fileext = ".csv.gz")
write.csv(mtcars_df, gzfile(csv_path), row.names = FALSE)

# DuckDB reads it directly
tbl_csv <- DuckDBTable(csv_path,
                       datacols = colnames(mtcars),
                       keycols = list(model = mtcars_df$model))
```

### From dplyr Connections

Advanced users can construct tables from existing dplyr connections:

```{r DuckDBTable-from-dplyr}
library(dplyr)

# Get the DuckDB connection
conn <- acquireDuckDBConn()

# Create a dplyr connection with filtering
tbl_conn <- tbl(conn, paste0("read_parquet('", mtcars_path, "')")) %>%
    filter(cyl >= 6)

# Wrap in DuckDBTable
tbl_filtered <- DuckDBTable(tbl_conn,
                            datacols = colnames(mtcars),
                            keycols = list(model = mtcars_df$model[mtcars$cyl >= 6]))
tbl_filtered
```


## The DuckDBTable Contract

Classes extending *DuckDBTable* or using it as a backend must understand its
contract:

### Required Operations

1. **Dimensionality**: `nrow()`, `ncol()`, `dim()` via keycols and datacols
2. **Subsetting**: `[` operator for extracting sub-tables
3. **Accessors**: `keynames()`, `keydimnames()`, `colnames()`
4. **Coercion**: `as.data.frame()` for materialization

### Key Dimension Semantics

- **0 key dimensions**: No row names, generates row numbers internally
- **1 key dimension**: Row names for DataFrames
- **2+ key dimensions**: Array indices for higher-dimensional structures

### Special Row Number Encoding

When no keycols are provided, *DuckDBTable* uses a special encoding:

```{r row-number-encoding}
# Table without explicit row names
tbl_no_keys <- DuckDBTable(mtcars_path, datacols = colnames(mtcars))

# Check if it has row numbers
.has_row_number(tbl_no_keys)

# The keycols encoding
.keycols(tbl_no_keys)
```

This encoding `c(NA_integer64_, -n)` indicates row numbers from 1 to n.


## Matrix Statistics on Tables

Because *DuckDBTable* is n-dimensional, it supports `r CRANpkg("matrixStats")`
operations when `nkey(x) >= 2`:

```{r DuckDBTable-matrixStats}
# Create a 2D table (matrix-like) in long format
state_path <- tempfile(fileext = ".parquet")
state_df <- data.frame(
  state = rep(rownames(state.x77), times = ncol(state.x77)),
  metric = rep(colnames(state.x77), each = nrow(state.x77)),
  value = as.vector(state.x77)
)
arrow::write_parquet(state_df, state_path)

tbl_2d <- DuckDBTable(state_path,
                      datacols = "value",
                      keycols = list(state = rownames(state.x77),
                                     metric = colnames(state.x77)))

# Matrix statistics work directly
colMeans(tbl_2d)
rowSums(tbl_2d)
```

This is a key architectural decision: *DuckDBTable* implements matrixStats
methods because it's fundamentally n-dimensional, not because it's trying to
be a matrix. This enables both *DuckDBDataFrame* and *DuckDBArray* to inherit
efficient statistical operations.


# The DuckDBDataFrame Class

## Class Structure

*DuckDBDataFrame* extends both *DuckDBTable* and *DataFrame*:

```{r DuckDBDataFrame-class}
showClass("DuckDBDataFrame")
```

This dual inheritance provides:
- *DuckDBTable*: SQL backend and n-dimensional operations
- *DataFrame*: Bioconductor DataFrame API and metadata support

### Validity Constraints

*DuckDBDataFrame* adds one constraint: `nkey(x) <= 1`. This ensures the object
is 2-dimensional (0 or 1 key dimensions).

```{r DuckDBDataFrame-validity, error=TRUE}
# This would fail: 2 key dimensions
# DuckDBDataFrame(tbl_2d)  # Error: keycols has more than one element
```


## Construction

The *DuckDBDataFrame* constructor wraps *DuckDBTable* construction:

```{r DuckDBDataFrame-construct}
df <- DuckDBDataFrame(mtcars_path,
                      datacols = colnames(mtcars),
                      keycol = list(model = mtcars_df$model))
df
```

Note the parameter name change: `keycol` (singular) instead of `keycols`,
emphasizing the single-dimension constraint.


## DataFrame Operations

*DuckDBDataFrame* supports the full *DataFrame* API from `r Biocpkg("S4Vectors")`:

### Column Access

```{r DuckDBDataFrame-columns}
# Extract single column (returns DuckDBColumn)
mpg <- df$mpg
mpg

# Extract multiple columns
df[, c("mpg", "cyl", "hp")]
```

### Row Subsetting

```{r DuckDBDataFrame-rows}
# By row names
df[c("Mazda RX4", "Honda Civic"), ]

# By logical vector
df[df$mpg > 25, ]

# By integer positions (order not preserved!)
df[1:5, ]
```

### Column Assignment

```{r DuckDBDataFrame-assign}
# Create a copy
df_copy <- df

# Add computed column
df_copy$efficiency <- df$mpg / df$hp
df_copy[1:3, c("mpg", "hp", "efficiency")]
```

### Combining

```{r DuckDBDataFrame-combine}
# Column binding
df_subset <- df[, c("mpg", "cyl")]
df_combined <- cbind(df_subset, hp = df$hp)
df_combined[1:3, ]
```


## Metadata Support

*DuckDBDataFrame* fully supports DataFrame metadata:

```{r DuckDBDataFrame-metadata}
# Column metadata
mcols(df) <- DataFrame(
    description = c("Miles/(US) gallon", "Number of cylinders",
                   "Displacement (cu.in.)", "Gross horsepower",
                   "Rear axle ratio", "Weight (1000 lbs)",
                   "1/4 mile time", "Engine (0 = V-shaped, 1 = straight)",
                   "Transmission (0 = automatic, 1 = manual)",
                   "Number of forward gears", "Number of carburetors"),
    row.names = colnames(mtcars)
)

mcols(df)
```


# The DuckDBColumn Class

## Class Structure

*DuckDBColumn* represents a single extracted column, wrapping a *DuckDBTable*
with `ncol(table) == 1`:

```{r DuckDBColumn-class}
showClass("DuckDBColumn")
```

### Slots

**table** (*DuckDBTable*)
: A single-column DuckDBTable containing the data.

**metadata** (`list`)
: Column metadata from the parent DataFrame's mcols.


## Vector-like Behavior

*DuckDBColumn* behaves like an ordinary R vector:

```{r DuckDBColumn-ops}
mpg <- df$mpg

# Length
length(mpg)

# Subsetting
mpg[1:5]

# Arithmetic
mpg / 2

# Comparison
mpg > 20

# Summary
mean(mpg)
median(mpg)
```

## Lazy Evaluation

Operations on *DuckDBColumn* remain lazy until values are needed:

```{r DuckDBColumn-lazy}
# This builds a SQL expression, doesn't execute
efficiency <- df$mpg / df$hp
efficiency

# Materialization happens on demand
as.vector(efficiency)[1:5]
```


# The DuckDBAtomicList Class

## Purpose

*DuckDBAtomicList* represents list columns where each element is an atomic
vector. In DuckDB, these are stored as `LIST[]` columns.

```{r DuckDBAtomicList-class}
showClass("DuckDBAtomicList")
```

## Construction

```{r DuckDBAtomicList-construct}
# Create data with list column
list_df <- data.frame(
    id = 1:3,
    name = c("Alice", "Bob", "Charlie"),
    scores = I(list(c(85, 90, 88), c(92, 87), c(78, 82, 85, 90)))
)

list_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(list_df, list_path)

# Create DuckDBDataFrame
df_list <- DuckDBDataFrame(list_path, keycol = "id")

# Extract list column
scores <- df_list$scores
scores
```

## List Operations

```{r DuckDBAtomicList-ops}
# Element lengths
elementNROWS(scores)

# Extract single element
scores[[1]]

# Unlisting
unlist(scores)
```


# The DuckDBEmbeddings Class

## Purpose

*DuckDBEmbeddings* represents fixed-length numeric arrays, typically used for
embeddings or feature vectors. In DuckDB, these are stored as `ARRAY[n]` columns.

```{r DuckDBEmbeddings-class}
showClass("DuckDBEmbeddings")
```

## Construction

```{r DuckDBEmbeddings-construct}
# Create data with fixed-length arrays
embeddings_df <- data.frame(
    id = 1:5,
    embedding = I(lapply(1:5, function(i) runif(10)))
)

emb_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(embeddings_df, emb_path)

# Create DuckDBDataFrame
df_emb <- DuckDBDataFrame(emb_path, keycol = "id")

# Extract embeddings column (returns DuckDBEmbeddings)
emb <- df_emb$embedding
emb
```

## Matrix-like Operations

*DuckDBEmbeddings* behaves like a matrix where each row is an embedding:

```{r DuckDBEmbeddings-ops}
# Dimensions
dim(emb)

# Subsetting
emb[1:3, 1:5]
```


# SQL Function Discovery

## The sql_fun() Function

`r Biocpkg("DuckDBDataFrame")` provides `sql_fun()` to discover available SQL
functions for a given data type:

```{r sql_fun}
# Find functions for numeric columns
sql_fun(df$mpg)
```

This queries DuckDB's function catalog to find compatible functions.


## The sql_call() Function

`sql_call()` applies any DuckDB SQL function to a DuckDBDataFrame object:

```{r sql_call}
# Apply SQL functions
sql_call(df$mpg, "sqrt")
sql_call(df$mpg, "round", 1)

# String functions
df_char <- DuckDBDataFrame(mtcars_path, datacols = "model", keycol = "model")
sql_call(df_char$model, "upper")
```

This provides a low-level interface to DuckDB's extensive SQL function library.


# Connection Management

## Shared Connection Pool

`r Biocpkg("DuckDBDataFrame")` manages a shared DuckDB connection:

```{r connection-mgmt}
# Acquire the shared connection
conn <- acquireDuckDBConn()
conn

# All DuckDBDataFrame objects use this connection
identical(dbconn(df), conn)

# Release when done (usually not needed)
# releaseDuckDBConn()
```

This design ensures:
- Single DuckDB process per R session
- Efficient resource usage
- Consistent transaction semantics


## Direct SQL Access

Advanced users can access the underlying connection for custom SQL:

```{r direct-sql}
library(DBI)

# Execute custom SQL
dbGetQuery(conn, "SELECT AVG(mpg) as avg_mpg FROM read_parquet(?)", 
           params = list(mtcars_path))
```


# Advanced Topics

## Type System

*DuckDBTable* automatically detects and handles DuckDB types:

```{r type-system}
# Get column types
coltypes(df)

# Cast types
df_copy <- df
coltypes(df_copy)[["mpg"]] <- "integer"
df_copy$mpg[1:5]
```

## Dimension Tables

For partitioned data, dimension tables map indices to grid groups:

```{r dimtbls}
# Create dimension tables (usually from createDimTables)
dimtbl <- DataFrame(
    state = rownames(state.x77),
    region = rep(c("West", "East"), length.out = nrow(state.x77))
)

# Attach to table
dimtbls(tbl_2d) <- list(state = dimtbl)

# Now filtering can use region
dimtbls(tbl_2d)
```


# Use Cases

## Large Metadata Tables

Single-cell experiments with millions of cells generate large metadata tables.
*DuckDBDataFrame* keeps this on disk:

```{r use-case-metadata, eval=FALSE}
# Hypothetical: 1 million cell metadata
cell_metadata <- DuckDBDataFrame("cells.parquet", keycol = "cell_id")

# Efficient filtering
cd4_cells <- cell_metadata[cell_metadata$cell_type == "CD4+ T cell", ]

# Aggregation
table(cell_metadata$cell_type)
```

## Data Integration

Combine data from multiple Parquet files:

```{r use-case-integration, eval=FALSE}
# Read multiple files as a single table
multi_path <- "experiment_*/cells.parquet"
all_cells <- DuckDBDataFrame(multi_path, keycol = "cell_id")
```

## SQL-Based Transformations

Leverage DuckDB's SQL engine for complex transformations:

```{r use-case-sql}
# Complex filtering with SQL
high_perf <- df[df$hp > 100 & df$mpg > 15, ]

# This translates to efficient SQL:
# SELECT * FROM table WHERE hp > 100 AND mpg > 15
```


# Integration with Bioconductor

## S4Vectors Compatibility

*DuckDBDataFrame* is a full *DataFrame* implementation:

```{r s4vectors-compat}
library(S4Vectors)

# Works with DataFrame functions
subset(df, mpg > 20, select = c("mpg", "cyl", "hp"))

# Combines with regular DataFrames
regular_df <- DataFrame(x = 1:3, y = letters[1:3])
# Can be used in mixed contexts
```

## Extension by Other Packages

`r Biocpkg("DuckDBArray")` and `r Biocpkg("DuckDBGRanges")` extend this
foundation:

- *DuckDBArray* uses *DuckDBTable* with `nkey >= 2` for arrays
- *DuckDBGRanges* uses *DuckDBDataFrame* for genomic coordinates
- Both inherit SQL translation and lazy evaluation


# Performance Considerations

## When to Use DuckDBDataFrame

**Advantages**:
- Data larger than RAM
- Columnar operations (aggregations, filtering)
- Parquet files already on disk
- SQL-based transformations

**Limitations**:
- Random access slower than in-memory
- Row-wise operations less efficient
- Overhead for small datasets


## Optimization Tips

1. **Use Parquet format**: Columnar storage with compression
2. **Filter early**: Push predicates to SQL
3. **Avoid materialization**: Keep operations lazy
4. **Use column selection**: Only read needed columns
5. **Leverage indexes**: DuckDB can use Parquet statistics


# Session Information

```{r sessioninfo}
sessionInfo()
```
