---
title: "Design and extension of DuckDBDataFrame"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r BiocStyle::doc_date()`; Modified: 6 July 2026"
package: DuckDBDataFrame
vignette: >
  %\VignetteIndexEntry{2. Design and extension of DuckDBDataFrame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      error = FALSE, warning = FALSE, message = FALSE)
```

# Scope

This vignette is for developers: it documents the `DuckDBTable` abstraction that
underlies `r Biocpkg("DuckDBDataFrame")`, how R operations become SQL, and how other
packages build on it. For day-to-day use see *Introduction to DuckDBDataFrame*.

```{r load}
library(DuckDBDataFrame)
```

# The classes

`r Biocpkg("DuckDBDataFrame")` extends the `r Biocpkg("S4Vectors")` framework with
DuckDB-backed versions of its core tabular classes:

| Class | Extends | Purpose |
|-------|---------|---------|
| `DuckDBTable` | `RectangularData` | N-dimensional table with a SQL backend |
| `DuckDBDataFrame` | `DataFrame` + `DuckDBTable` | 2-D tabular data with row/column names |
| `DuckDBColumn` | `Vector` | a single extracted (atomic) column |
| `DuckDBAtomicList` | `List` | list columns (DuckDB `LIST[]`) |
| `DuckDBEmbeddings` | matrix-like | fixed-length arrays (DuckDB `ARRAY[n]`) |

`DuckDBTable` is the foundation. `DuckDBDataFrame` is the 2-D case (it adds the
constraint `nkey(x) <= 1`); the same `DuckDBTable` with two or more key dimensions is
what `r Biocpkg("DuckDBArray")` wraps for arrays.

# The DuckDBTable abstraction

A `DuckDBTable` records a *query*, not data. Its slots are:

- **`conn`** --- a `tbl_duckdb_connection` (dplyr/dbplyr) over a DuckDB relation:
  Parquet/CSV files or an in-memory table.
- **`datacols`** --- a named `expression`; each element is a column reference
  (`as.name("mpg")`) or a computation (`call("/", as.name("mpg"), as.name("hp"))`)
  that translates to SQL.
- **`keycols`** --- a named `list` of dimension index vectors: length 0 (row numbers
  generated internally), 1 (row names, a `DuckDBDataFrame`), or ≥2 (array indices).
- **`dimtbls`** --- an optional locked environment of dimension lookup tables for
  partition pruning.

Operations build up `datacols`/`keycols` and the `conn` query lazily; materialization
(`as.data.frame()`, `as.vector()`) is deferred until values are needed. This is what
lets a `DuckDBTable` describe data larger than memory and push filters and arithmetic
into DuckDB's columnar engine (predicate pushdown, direct Parquet scans).

## Construction

`DuckDBTable()` accepts a Parquet or CSV path, or an existing dplyr connection:

```{r construct}
mtcars_df <- cbind(model = rownames(mtcars), mtcars)
path <- tempfile(fileext = ".parquet")
arrow::write_parquet(mtcars_df, path)

tbl <- DuckDBTable(path, datacols = colnames(mtcars),
                   keycols = list(model = mtcars_df$model))
dim(tbl)
```

## The contract

Classes that extend `DuckDBTable` or use it as a backend rely on:
`nrow()`/`ncol()`/`dim()` (from `keycols`/`datacols`), `[` for sub-tables,
`keynames()`/`keydimnames()`/`colnames()` accessors, and `as.data.frame()` for
materialization.

## Key-dimension semantics

The number of key dimensions determines the shape: 0 → no row names (row numbers
generated), 1 → a `DuckDBDataFrame`, ≥2 → an array. With no `keycols`, `DuckDBTable`
uses a compact row-number encoding:

```{r rownum}
tbl0 <- DuckDBTable(path, datacols = colnames(mtcars))
.has_row_number(tbl0)
```

# From R to SQL

Column operations become SQL, at the table level. A `DuckDBColumn` computation like
`df$mpg / df$hp` records `call("/", as.name("mpg"), as.name("hp"))` in `datacols`;
row/column summaries on arrays (via `r Biocpkg("DuckDBArray")`) become `GROUP BY`
aggregations. The `sql_fun()` / `sql_call()` helpers expose DuckDB's function catalog
so any SQL scalar function can be applied without leaving R.

# Connection management

`r Biocpkg("DuckDBDataFrame")` keeps a single shared DuckDB connection per session,
acquired lazily:

```{r conn}
conn <- acquireDuckDBConn()
identical(dbconn(tbl), conn)
```

One process per session gives consistent semantics and efficient resource use;
`releaseDuckDBConn()` tears it down (rarely needed). The connection also configures a
writable extension directory so extension install/load works on shared or read-only
R libraries. Advanced callers can run arbitrary SQL through `dbconn()` with
`r CRANpkg("DBI")`.

# Dimension tables

For partitioned data, dimension tables map index values to partition groups so a
query can prune to the relevant Parquet partitions:

```{r dimtbls}
state_df <- data.frame(
    state = rep(rownames(state.x77), times = ncol(state.x77)),
    metric = rep(colnames(state.x77), each = nrow(state.x77)),
    value = as.vector(state.x77))
sp <- tempfile(fileext = ".parquet"); arrow::write_parquet(state_df, sp)
tbl2 <- DuckDBTable(sp, datacols = "value",
                    keycols = list(state = rownames(state.x77),
                                   metric = colnames(state.x77)))
dimtbls(tbl2) <- list(state = DataFrame(
    row.names = rownames(state.x77),
    region = rep(c("West", "East"), length.out = nrow(state.x77))))
names(dimtbls(tbl2))
```

# Extending DuckDBDataFrame

Other suite packages build directly on this foundation:

- `r Biocpkg("DuckDBArray")` wraps a `DuckDBTable` with `nkey >= 2` as a
  `DelayedArray` backend (see that package's *Implementing the DuckDBArray backend*
  vignette).
- `r Biocpkg("DuckDBGRanges")` uses a `DuckDBDataFrame` to hold genomic coordinates.

Both inherit the SQL translation and lazy evaluation described here, so a new
backend mostly needs to define how its data maps onto `keycols`/`datacols` and which
operations to push into SQL.

# Session information

```{r sessioninfo}
sessionInfo()
```
